Persistent somatic symptoms are key to individual illness perception at one year after COVID-19 in a cross-sectional prospective cohort study

### 2023-01-09 paper/markdown/manuscript.Rmd
Revision

---
title: "Persistent somatic symptoms are key to individual illness perception at one year after COVID-19"
subtitle: "Manuscript file"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

author: 
  - Katharina Huefner:
      institute: [psych]
  - Piotr Tymoszuk:
      institute: [daas]
  - Sabina Sahanic:
      institute: [int2]
  - Anna Luger:
      institute: [radio]
  - Anna Boehm:
      institute: [int2]
  - Alex Pizzini:
      institute: [int2]
  - Christoph Schwabl:
      institute: [radio]
  - Sabine Koppelstätter:
      institute: [int2]
  - Katharina Kurz:
      institute: [int2]
  - Malte Asshoff:
      institute: [int2]
  - Birgit Mosheimer-Feistritzer:
      institute: [int2]
  - Maximilian Coen:
      institute: [int2]
  - Bernhard Pfeifer:
      institute: [umit]
  - Verena Raas:
      institute: [neuro]
  - Alexander Egger:
      institute: [zimcl]
  - Gregor Hörmann:
      institute: [zimcl]
  - Raimund Helbok:
      institute: [neuro]
  - Ewald Wöll:
      institute: [zams]
  - Günter Weiss:
      institute: [int2]
  - Gerlig Widmann:
      institute: [radio]
  - Thomas Sonnweber:
      institute: [int2]
  - Ivan Tancevski:
      institute: [int2]
  - Barbara Sperner-Unterweger:
      institute: [psych]
  - Judith Löffler-Ragg:
      institute: [int2]
      email: judith.loeffler@i-med.ac.at
      correspondence: true

institute:
  - psych: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Innsbruck, Austria
  - daas: Data Analytics As a Service Tirol, Innsbruck, Austria
  - int2: Department of Internal Medicine II, Medical University of Innsbruck, Innsbruck Austria
  - radio: Department of Radiology, Medical University of Innsbruck, Innsbruck, Austria
  - umit: Division for Health Networking and Telehealth, Biomedical Informatics and Mechatronics, UMIT, Hall in Tyrol, Austria
  - neuro: Department of Neurology, Medical University of Innsbruck, Innsbruck, Austria
  - zimcl: Central Institute of Medical and Chemical Laboratory Diagnostics, University Hospital Innsbruck, Innsbruck, Austria
  - zams: Department of Internal Medicine, St. Vinzenz Hospital, Zams, Austria

@@

csl: frontiers-in-immunology.csl

@@

__Background:__
Sequelae of Coronavirus disease 2019 (COVID-19) were investigated by both patient-initiated and academic initiatives. 
Patient´s subjective illness perceptions might differ from physician’s clinical assessment results. 
Herein, we explored factors influencing patient’s perception during COVID-19 recovery.

__Methods:__
Participants of the prospective observation CovILD study with persistent somatic symptoms or cardiopulmonary findings at the clinical follow-up one year after COVID-19 were analyzed (n = `r nrow(cov_data$clear_data)`). 
Explanatory variables included baseline demographic and comorbidity data, COVID-19 course and one-year follow-up data of persistent somatic symptoms, physical performance, lung function testing (LFT), chest computed tomography (CT) and trans-thoracic echocardiography (TTE). 
Factors affecting illness perception (Brief Illness Perception Questionnaire, BIPQ) were identified by penalized multi-parameter regression and unsupervised clustering.

__Results:__
In modeling, `r (blass_mod$cv_fit_stats$ipq_total$estimate[4] %>% signif(2) %>% unname) * 100`% of overall illness perception variance at one year after COVID-19 was attributed to fatigue intensity, reduced physical performance, hair loss and  baseline respiratory comorbidity. 
Overall illness perception was independent of LFT results, pulmonary lesions in CT or heart abnormality in TTE. 
As identified by clustering, persistent somatic symptom count, fatigue, diminished physical performance, dyspnea, hair loss and sleep problems at the one-year follow-up and severe acute COVID-19 were associated with the BIPQ domains of concern, emotional representation, complaints, disease timeline and consequences.

__Conclusion:__
Persistent somatic symptoms rather than clinical assessment results, revealing lung and heart abnormalities, impact on severity and quality of illness perception at one year after COVID-19 and may foster unhelpful coping mechanisms. 
Besides  COVID-19 severity, individual illness perception should be taken into account when allocating rehabilitation and psychological therapy resources.

__Study registration:__ 
ClinicalTrials.gov: NCT04416100.

@@

A sizable fraction of coronavirus disease 2019 (COVID-19) patients is affected by protracted somatic symptoms, cardiopulmonary pathology and mental health disorders [@Sonnweber2020; @Sahanic2021; @Luger2022; @Sonnweber2022; @Hufner2022; @Al-Aly2021; @Davis2021; @Soriano2022; @Evans2021a]. 
Persistent COVID-19-related symptoms have been initially described by patient and social media initiatives [@McCorkell2021] and subsequently, recognized as the 'post COVID-19 condition' by the clinical community [@Sonnweber2020; @Sahanic2021; @Davis2021; @Soriano2022; @Al-Aly2021].
Yet, the patients' and clinicians' characteristics were not always consistent, which was observed also in other conditions, such as chronic obstructive pulmonary disease [@Miravitlles2013; @Schwartz2020] or functional disorders [@Schumacher2013].
For post COVID-19 condition, the matter is further complicated by a broad range of patient-reported manifestations and widespread character of the disease [@Davis2021; @Sahanic2021].
Since it is becoming increasingly evident that many COVID-19 convalescents experience prolonged severe individual suffering [@Hufner2022; @Huang2021; @Davis2021; @Evans2021a], and pose a significant healthcare and socioeconomic challenge [@Al-Aly2021], it is critically important to characterize and understand individual illness perception following COVID-19.

Illness perceptions can be divided into cognitive or emotional components [@Broadbent2006]. 
The cognitive representations include self-perceived consequences, expected duration, personal control, expected effect of treatment, symptom perception and understanding of the disease. 
The emotional components encompass patient's concerns and emotions e.g. fear, anger or distress associated with the disease [@Broadbent2006; @Weinman1996].
According to the common-sense model of self regulation (CSM), a theoretical framework enabling understanding of how people cope with threats to their health, such illness perceptions are influenced by situational stimuli such as symptoms, health information and patient's knowledge [@Man2021; @Leventhal2016; @Diefenbach1996]. 
The individual illness perception was shown to influence disease coping, adjustment to adverse life events or chronic conditions and compliance with prevention, treatment and rehabilitation [@Hagger2010; @Figueiras2019], also in context of acute COVID-19 [@DiasNeto2021; @Lugo-Gonzalez2020; @Vollmann2021].
Finally, severe illness perception including the emotional response, concerns and consequence components was found associated with anxiety, depression and stress both in the general population and in COVID-19 patients [@Wierenga2021; @Sadeghian2022; @Hamama2022; @Burke2020].

Herein, we investigated severity of overall illness perception and the illness perception components in a cross-sectional COVID-19 convalescent collective [@Sonnweber2020; @Sahanic2021; @Luger2022; @Sonnweber2022] with incomplete somatic symptom resolution or COVID-19-related lung and heart pathology in the follow-up clinical assessment at one year after diagnosis.

@@

Participants of the longitudinal observation CovILD study (ClinicalTrials.gov: NCT04416100) were recruited between April and June 2020 at three Austrian clinical centers [@Sonnweber2020; @Luger2022; @Sonnweber2022]. 
The study inclusion criteria were age $\geq$ 18 years and symptomatic PCR-confirmed SARS-CoV-2 infection. 
The follow-up visits were scheduled at two, three, six months and one year after COVID-19 diagnosis. 
A subset of participants (n = `r nrow(cov_data$clear_data)`) was analyzed who completed the Brief Illness Perception Questionnaire (BIPQ) [@Broadbent2006] and displayed (1) COVID-19-related persistent somatic symptoms or (2) any abnormality in chest computed tomography (CT) or (4) any lung function testing (LFT) deficits or (5) any heart abnormality in trans-thoracic echocardiography (TTE) at the one-year follow-up (__Figure \@ref(fig:fig-consort)__).

The study was performed in accordance with the Declaration of Helsinki and the European Data Policy. All participants gave written informed consent. 
The study protocol was approved by the ethics committee at the Medical University of Innsbruck (approval number 1103/2020).

@@

For full descriptions of procedures and variables, see __Supplementary Methods__ and __Supplementary Table S1__.

Baseline clinical and acute COVID-19 data were recorded retrospectively at the two-month follow-up based on the patient's interview and electronic patient records [@Sonnweber2020]. 
Study participants were classified as ambulatory (outpatient, WHO grade 1 - 2), moderate (hospitalized, without oxygen therapy or mask/nasal prongs oxygen, WHO 3 - 4) and severe COVID-19 survivors (hospitalized with non-invasive ventilation or high-flow oxygen or mechanical ventilation, WHO 5 - 7).

Physical performance was rated with the Eastern Cooperative Oncology Group scale (ECOG). 
Dyspnea was scored with the Modified Medical British Research Council scale (mMRC).
Fatigue at one-year follow-up was rated with likert and bimodal Chalder's Fatigue Scales (CFS) [@Chalder1993; @Morriss1998]. 
Exertional capacity at the one-year follow-up was assessed by six-minute walking distance (SMWD) test and compared with the reference values [@Crapo2012].

The following COVID-19-related persistent somatic symptoms at the one-year follow-up were analyzed: 
reduced physical performance (ECOG $\geq$ 1), 
dyspnea (mMRC $\geq$ 1), 
self-reported cough (yes/no item), 
self-reported sleep problems (yes/no), 
self-reported night sweating (yes/no), 
self-reported hyposmia or anosmia (yes/no), 
self-reported dermatological symptoms (yes/no), 
self-reported gastrointestinal symptoms (yes/no), 
self-reported hair loss (yes/no), 
significant fatigue (bimodal CFS $\geq$ 4).

@@

Illness perception was investigated with the 8-item BIPQ (Q1 - Q8) [@Broadbent2006]. 
Each item was rated with an 11-point likert scale. 
The illness perception score was defined as a sum of all BIPQ items, with the negative items Q3, Q4 and Q7 inverted.

@@

The primary analysis endpoints were illness perception score values and its influencing factors among demographic, clinical and somatic symptom data at the one-year after COVID-19 in convalescents affected by persistent somatic symptoms or residual cardiopulmonary findings.
The secondary analysis endpoints were particular BIPQ component scores and their influencing factors.

@@

Details of statistical analysis are provided in __Supplementary Methods__. 
Data analysis was done with R version 4.2.0 (R Foundation for Statistical Computing). 
Differences in categorical variable distribution were assessed by $\chi^2$ test with Cramer V effect size statistic. 
Statistical significance for numerical variables was investigated by Mann-Whitney test with Wilcoxon r effect size statistic or Kruskal-Wallis test with $\eta^2$ effect size statistic [@Kassambara2021]. 
Correlations were investigated by Spearman's test.

Multi-parameter modeling was done with the Elastic Net (package _glmnet_) [@Zou2005; @Friedman2010], LASSO (least absolute shrinkage and selection operator; package _glmnet_) [@Tibshirani1996; @Friedman2010] and Bayesian LASSO (package _monomvn_) [@Park2012; @Gramacy2022] algorithms. 
The response, illness perception score, was square-root transformed to guarantee normality. 
Both first and second order terms of numeric explanatory variables were included in the models. 
Numeric explanatory and response variables were Z-score normalized. 
The optimal $\lambda$ for Elastic Net and LASSO were obtained by 200-repeats 10-fold cross-validation (CV). 
The 'sparsity' parameter in Bayesian LASSO was found by 10-repeats 10-fold cross-validation (package _caret_) [@Kuhn2008]. 
Explained variance ($R^2$) and model root mean squared error (RMSE) were assessed in the entire data set and 10-repeats 10-fold CV (package _caret_) [@Kuhn2008]. 
Elastic Net and LASSO coefficients were calculated for the optimal $\lambda$ values. 
Bayesian LASSO coefficients were calculated as medians over all algorithm iterations [@Park2012].
Variables with non-zero coefficients in all three models were deemed key factors for the illness perception score.

Clustering by the BIPQ items was accomplished with the PAM algorithm (partitioning around medoids; Euclidean distance) [@Schubert2019; @Drost2018].

@@

The CovILD cohort was recruited between April and June 2020. 
Out of 145 participants enrolled, `r nrow(cov_data$clear_data)` individuals with COVID-19-related persistent somatic symptoms, lung CT, LFT or cardiological abnormalities and complete BIPQ were included in the analysis. 
The major reasons of patient dropout were missing follow-up visits and incomplete BIPQ (__Figure \@ref(fig:fig-consort)__).
The participants were predominantly male (`r get_percent(cov_data$clear_data, "sex")['male']`%), the median age at COVID-19 diagnosis was `r get_median_iqr(cov_data$clear_data, "age")['median']` years (IQR: `r get_median_iqr(cov_data$clear_data, "age")['iqr']`), over one-third were active or ex-smokers (`r get_percent(cov_data$clear_data, "smoking_history")['yes']`%). 
Most participants suffered from comorbidities (`r get_percent(cov_data$clear_data, "comorb_present")['yes']`%), with cardiovascular disease, metabolic and respiratory disorders as the leading conditions. 
The participants were classified by COVID-19 severity as ambulatory (`r get_percent(cov_data$clear_data, "cat_WHO")[1]`%), moderate (hospitalized, no intensive care , no oxygen therapy, `r get_percent(cov_data$clear_data, "cat_WHO")[2]`%) and severe COVID-19 survivors (hospitalized, oxygen therapy or intensive care, `r get_percent(cov_data$clear_data, "cat_WHO")[3]`%). 
The ambulatory COVID-19 subset had the lowest median age, rate of smokers and comorbidity frequency (__Table \@ref(tab:tab-cohort-baseline)__).

>>>>

---
title: "Persistent somatic symptoms are key to individual illness perception at one year after COVID-19 in a cross-sectional prospective cohort study"
subtitle: "Illness perception one year after COVID-19"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

author: 
  - Katharina Hüfner, MD:
      institute: [psych]
  - Piotr Tymoszuk, PhD:
      institute: [daas]
  - Sabina Sahanic, MD:
      institute: [int2]
  - Anna Luger, MD:
      institute: [radio]
  - Anna Boehm, MD:
      institute: [int2]
  - Alex Pizzini, MD:
      institute: [int2]
  - Christoph Schwabl, MD:
      institute: [radio]
  - Sabine Koppelstätter, MD:
      institute: [int2]
  - Katharina Kurz, MD:
      institute: [int2]
  - Malte Asshoff, MD:
      institute: [int2]
  - Birgit Mosheimer-Feistritzer, MD:
      institute: [int2]
  - Bernhard Pfeifer, PhD:
      institute: [umit]
  - Verena Rass, MD PhD:
      institute: [neuro]
  - Andrea Schroll, MD PhD:
      institute: [int2]
  - Sarah Iglseder, MD:
      institute: [neuro]
  - Alexander Egger, MD:
      institute: [zimcl]
  - Ewald Wöll, MD:
      institute: [zams]
  - Günter Weiss, MD:
      institute: [int2]
  - Raimund Helbok, MD:
      institute: [neuro]
  - Gerlig Widmann, MD:
      institute: [radio]
  - Thomas Sonnweber, MD:
      institute: [int2]
  - Ivan Tancevski, MD:
      institute: [int2]
  - Barbara Sperner-Unterweger, MD:
      institute: [psych]
  - Judith Löffler-Ragg, MD:
      institute: [int2]
      email: judith.loeffler@i-med.ac.at
      phone: +43 (0)512/504-23251
      fax: +43 (0)512/504-23317
      correspondence: true

institute:
  - psych: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Innsbruck, Austria
  - daas: Data Analytics As a Service Tirol, Innsbruck, Austria
  - int2: Department of Internal Medicine II, Medical University of Innsbruck, Innsbruck, Austria
  - radio: Department of Radiology, Medical University of Innsbruck, Innsbruck, Austria
  - umit: Division for Health Networking and Telehealth, Biomedical Informatics and Mechatronics, UMIT, Hall in Tyrol, Austria
  - neuro: Department of Neurology, Medical University of Innsbruck, Innsbruck, Austria
  - zimcl: Central Institute of Medical and Chemical Laboratory Diagnostics, University Hospital Innsbruck, Innsbruck, Austria
  - zams: Department of Internal Medicine, St. Vinzenz Hospital, Zams, Austria

@@

csl: jpr.csl

@@

__Objective:__
Subjective illness perception (IP) can differ from physician’s clinical assessment results. 
Herein, we explored patient’s IP during <span custom-style = 'revision'>coronavirus disease 2019 (COVID-19)</span> recovery.

__Methods:__
Participants of the prospective observation CovILD study <span custom-style = 'revision'>(ClinicalTrials.gov: NCT04416100)</span> with persistent somatic symptoms or cardiopulmonary findings one year after COVID-19 were analyzed (n = `r nrow(cov_data$clear_data)`). 
Explanatory variables included demographic and comorbidity, COVID-19 course and one-year follow-up data of persistent somatic symptoms, physical performance, lung function testing, chest computed tomography and trans-thoracic echocardiography. 
Factors affecting IP (Brief Illness Perception Questionnaire) <span custom-style = 'revision'>one year after COVID-19</span> were identified by regularized modeling and unsupervised clustering.

__Results:__
In modeling, 
<span custom-style = 'revision'>`r 100 * (lasso_mod$cv_fit_stats$ipq_total$estimate[4] %>% signif(2) %>% unname)`%</span> 
of overall IP variance <span custom-style = 'revision'>(R^2^)</span> was attributed to fatigue intensity, reduced physical performance <span custom-style = 'revision'>and persistent somatic symptom count.</span> 
Overall IP was <span custom-style = 'revision'>largely</span> independent of <span custom-style = 'revision'>lung and heart findings revealed by imaging and function testing.</span> 
<span custom-style = 'revision'>In<span> clustering, 
persistent somatic symptom count <span custom-style = 'revision'>(Kruskal-Wallis test: `r clust_chara$test_results$fup %>% filter(variable == 'sympt_number') %>% .$plot_cap`)</span>, 
fatigue <span custom-style>(`r clust_chara$test_results$fup %>% filter(variable == 'Chalder_FS') %>% .$plot_cap`)</span>, 
diminished physical performance <span custom-style = 'revision'>($\chi^2$ test, Cramer V effect size statistic: `r clust_chara$test_results$fup %>% filter(variable == 'fatigue_sympt') %>% .$plot_cap`)<span>, 
dyspnea <span custom-style = 'revision'>(`r clust_chara$test_results$fup %>% filter(variable == 'dyspnoe_sympt') %>% .$plot_cap`)</span>, 
hair loss <span custom-style = 'revision'>(`r clust_chara$test_results$fup %>% filter(variable == 'hair_loss_sympt') %>% .$plot_cap`)</span>
and sleep problems <span custom-style = 'revision'>(`r clust_chara$test_results$fup %>% filter(variable == 'sleep_sympt') %>% .$plot_cap`)</span> 
were <span custom-style = 'revision'>strongly</span> associated with the concern, emotional representation, complaints, disease timeline and consequences <span custom-style = 'revision'>IP dimensions.</span>

__Conclusion:__
Persistent somatic symptoms rather than <span custom-style = 'revision'>abnormalities in cardiopulmonary testing influence IP</span> one year after COVID-19. 
<span custom-style = 'revision'>Modifying IP represents a promising innovative approach to treatment of post-COVID-19 condition. Besides COVID-19 severity, individual IP should guide rehabilitation and psychological therapy decisions.</span>

# Keywords

COVID-19, clustering, illness perception, persistent somatic symptoms, post-COVID-19 condition, regularized regression

@@

A subset of coronavirus disease 2019 (COVID-19) patients is affected by protracted somatic symptoms, cardiopulmonary pathology and mental health disorders [@Sonnweber2020; @Sahanic2021; @Luger2022; @Sonnweber2022; @Hufner2022; @Al-Aly2021; @Davis2021; @Soriano2022; @Evans2021a; @Sudre2021; @Chen2022]. 
Persistent COVID-19-related symptoms have been initially described by patient initiatives [@McCorkell2021] and subsequently recognized as 'post-COVID-19 condition' by the clinical community [@Sonnweber2020; @Sahanic2021; @Davis2021; @Soriano2022; @Al-Aly2021; @Sudre2021]. 
<span custom-style = 'revision'>Most patients recover completely from acute COVID-19 symptoms within four weeks. 
The clearance rate is particularly fast for non-specific infection complaints [@Sudre2021; @Sahanic2021; @Rass2022]. 
In a recent metaanalysis of 1.7 million patients, the pooled prevalence estimate of post-COVID-19 condition, i.e. symptoms present for 3 months or longer, was 43% [@Soriano2022; @Chen2022]. 
Yet, in a rigorously controlled study, approximately 12.7% of SARS-CoV-2-positive individuals suffered from persistent somatic symptoms directly associated with COVID-19 [@Ballering2022].
Fatigue, concentration and memory problems, sleep disorders, physical performance deficits, smell/taste disorders and hair loss, dyspnea and pain are the most common, frequently overlapping and relapsing manifestations [@Chen2022; @Sonnweber2022; @Sahanic2021; @Davis2021; @Rass2022; @Evans2021a; @Huang2021; @Ballering2022].</span>

Yet, patients' and clinicians' perceptions 
<span custom-style = 'revision'>of post-COVID-19 condition <span custom-style = 'revision'>and its impact on patients´ lives</span> are not always consistent, as observed also in chronic obstructive pulmonary disease [@Miravitlles2013; @Schwartz2020] or functional disorders [@Schumacher2013]. 
More specifically, persistent symptoms in many COVID-19 patients were not accompanied by lung, heart, biochemical, inflammatory and immunity abnormalities [@Sonnweber2022; @Sneller2022]. 
<span custom-style = 'revision'>Persistent somatic symptoms (PSS) are defined as any subjectively distressing somatic complaints, including medically explained, unexplained and functional symptoms [@Lehmann2021].</span>
<span custom-style = 'revision'>Manifestations of post-COVID-19 condition</span> like fatigue or neuro-cognitive disorders were also associated with depression, anxiety, stress and other signs of mental health impairment [@Al-Aly2021; @Hufner2022; @Sneller2022; @Bierbauer2022; @Wang2022; @Kachaner2022]. 
Intriguingly, somatic complaints comparable to post-COVID-19 condition symptoms were also observed in COVID-19-free individuals [@Wisk2022; @Behnood2022; @Matta2022] suggestive of effects of socioeconomic background or altered social behavior during the SARS-CoV-2 pandemic [@Man2021; @Tutzer2021]. 
Fatigue, sleep problems and dyspnea were also widespread in the general pre-pandemic population [@Hinz2017].</span>

Illness perception (IP) consists of cognitive and emotional dimensions [@Broadbent2006]. 
The cognitive dimensions comprise of self-perceived consequences, expected duration, personal control, expected treatment effect, symptom perception and understanding of the disease. 
The emotional dimensions encompass disease-associated concerns and emotions e.g. fear, anger or distress [@Broadbent2006; @Weinman1996].
According to the common-sense model of self regulation, aiming to explain coping with health threats, IP is influenced by situational stimuli such as symptoms, health information and patient's knowledge [@Man2021; @Leventhal2016; @Diefenbach1996]. 
IP was shown to influence adjustment to adverse life events or chronic conditions and compliance with prevention, treatment and rehabilitation [@Hagger2010; @Figueiras2019], also in COVID-19 [@DiasNeto2021; @Lugo-Gonzalez2020; @Vollmann2021].
Emotional response, concerns and consequence IP dimensions correlate with anxiety, depression and stress both in the general population and acute COVID-19 patients [@Wierenga2021; @Sadeghian2022; @Hamama2022; @Burke2020; @Chen2022].
<span custom-style = 'revision'>Similar observations were made recently in individuals with post-COVID-19 condition recruited from self-help and social media platforms [@Bierbauer2022]. 
However, characterization of IP and its influencing factors in a defined observational COVID-19 cohort is still missing. 
In particular, it is unclear how PSS, lung and heart pathology [@Sonnweber2020; @Sonnweber2022; @Luger2022] shape IP during long-term COVID-19 recovery.</span>

<span custom-style = 'revision'>We aimed to describe IP and IP components, identify demographic and clinical variables affecting IP and explore IP patterns in convalescents suffering from persistent somatic symptoms (PSS), lung or heart abnormalities one year after COVID-19.
To this end, we gauged IP in a long-term observatory cross-sectional COVID-19 collective [@Sonnweber2020; @Sahanic2021; @Luger2022; @Sonnweber2022] with the Brief Illness Perception Questionnaire (BIPQ) [@Broadbent2006] and investigated components, influential factors and patterns of IP by factor analysis, regularized multi-parameter modeling and unsupervised PAM clustering.</span>

@@

Participants of the longitudinal observation CovILD study (ClinicalTrials.gov: NCT04416100) were recruited <span custom-style = 'revision'>during the first wild-type SARS-CoV-2 outbreak</span> between April and June 2020 at <span custom-style = 'revision'>the University Hospital of Innsbruck, the St. Vinzenz Hospital in Zams and the Karl-Landsteiner Rehabilitation Facility in Münster (all in Austria) [@Sonnweber2020; @Luger2022; @Sonnweber2022].</span>
The study inclusion criteria were age $\geq$ 18 years and symptomatic, PCR-confirmed SARS-CoV-2 infection. 
The follow-ups were scheduled at two, three, six months and one year after COVID-19 diagnosis.

<span custom-style = 'revision'>In total, 190 individuals were screened for participation. 
The screened subset size was limited by the number of COVID-19 cases diagnosed at the study centers, their follow-up availability (residency in the study region) and available resources for the follow-up assessments. 
No statistical sample size estimation was performed, as the kinetic of COVID-19 recovery was unknown at the pandemic onset.
Out of 145 enrolled individuals, `r nrow(cov_data$raw_data)` completed the one-year follow-up.</span>
A subset of participants (n = `r nrow(cov_data$clear_data)`) was analyzed who displayed (1) PSS or (2) any abnormality in chest computed tomography (CT) or (3) any lung function testing (LFT) deficits or (4) any heart abnormality in trans-thoracic echocardiography (TTE) at the one-year follow-up <span custom-style = 'revision'>and had a complete set of BIPQ, demographic and clinical analysis variables (__Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1 - S2__).</span>
<span custom-style = 'revision'>We decided to restrict the analysis to incomplete COVID-19 recovery, because BIPQ was initially designed for and validated in somatic conditions [@Broadbent2006].</span>

The study was performed in accordance with the Declaration of Helsinki and European Data Policy. 
All participants gave written informed consent. 
The study protocol was approved by the ethics committee at the Medical University of Innsbruck (approval number 1103/2020).

@@

For full descriptions of procedures and variables, see __Supplementary Methods__ and __Supplementary Tables S1__ and __S2__.

Baseline clinical and acute COVID-19 data were recorded retrospectively at the two-month follow-up [@Sonnweber2020]. 
Study participants were classified as ambulatory (outpatient, WHO <span custom-style = 'revision'>ordinal scale for clinical improvement</span> 1 - 2), moderate (hospitalized without oxygen therapy, WHO 3 - 4) and severe COVID-19 survivors (hospitalized with non-invasive ventilation or high-flow oxygen or mechanical ventilation, WHO 5 - 7).

Physical performance was rated with the Eastern Cooperative Oncology Group scale (ECOG). 
Dyspnea was scored with the Modified Medical British Research Council scale (mMRC).
Fatigue was rated with likert and bimodal Chalder's Fatigue Scales (CFS) [@Chalder1993; @Morriss1998]. 
Exertional capacity was assessed by six-minute walking distance (SMWD) [@Crapo2012].

The following, <span custom-style = 'revision'>PSS,</span> were analyzed at the one-year follow-up: 
reduced physical performance (ECOG $\geq$ 1), 
dyspnea (mMRC $\geq$ 1), 
self-reported cough (yes/no item), 
self-reported sleep problems (yes/no), 
self-reported night sweating (yes/no), 
self-reported hyposmia or anosmia (yes/no), 
self-reported dermatological symptoms (yes/no), 
self-reported gastrointestinal symptoms (yes/no), 
self-reported hair loss (yes/no), 
significant fatigue (bimodal CFS $\geq$ 4). 
<span custom-style = 'revision'>Records of reduced physical performance, dyspnea, cough, sleep problems, night sweating, hyposmia/anosmia and gastrointestinal symptoms were available for acute COVID-19 and all follow-ups [@Sonnweber2022].</span>

@@

IP was investigated with <span custom-style = 'revision'>a German translation of BIPQ [@Broadbent2006] with the following 8 items:</span> 
<span custom-style = 'revision'>Q1 consequences (0: no consequences, 10: very severe consequences), 
Q2 timeline (0: very short, 10: forever), 
Q3 lacking personal control (negative item, 0: full control, 10: no control at all), 
Q4 lacking treatment control (negative item, 0: full control, 10: no control at all), 
Q5 identity (0: no complaints, 10: multiple severe complaints), 
Q6 concerns (0: not at all, 10: extreme concerns), 
Q7 lacking coherence (negative item, 0: very good understanding, 10: no understanding at all) and 
Q8 emotional representation (0: no emotions at all, 10: extreme emotions).</span> 
Items were rated with 11-point likert scales, <span custom-style = 'revision'>with 0 corresponding to neutral or positive and 10 representing severely distorted or negative IP dimension. 
The Q3, Q4 and Q7 conceptualized as negative items were re-coded prior to analysis.</span>
<span custom-style = 'revision'>The total IP score was defined as all BIPQ item sum as a measure of overall negative IP alterations (0: neutral - 80: extremely severe) [@Broadbent2006].</span> 
<span custom-style = 'revision'>The emotion/concern/consequences score was defined as the Q1, Q2, Q5, Q6 and Q8 item sum (0: neutral - 50: extremely severe) and the lacking control/coherence score was defined as the Q3, Q4 and Q7 item sum (range: 0: good control/understanding - 30: no control/understanding at all) to measure intensity of the two main IP components identified by factor analysis.</span>

@@

<span custom-style = 'revision'>The primary analysis endpoints were demographic and clinical factors affecting total IP and its emotion/concern/consequences and lacking control/coherence components in COVID-19 convalescents with PSS or cardiopulmonary abnormalities at the one-year follow-up.</span>
<span custom-style = 'revision'>The secondary analysis endpoints were components of IP, patterns of IP dimensions explored by clustering as well as demographic and clinical factors differentiating between the IP clusters.</span>

@@

Statistical analysis details are provided in __Supplementary Methods__. 
Analysis was done with R version 4.2.0. 
<span custom-style = 'revision'>Numeric variables were presented as medians with interquartile ranges. 
Categorical variables were shown as percentages and counts.</span> 
Differences in categorical variable distribution were assessed by $\chi^2$ test with Cramer V effect size statistic. 
Statistical significance for numerical variables was investigated by Mann-Whitney test with r effect size statistic or Kruskal-Wallis test with $\eta^2$ effect size statistic. 
Correlations were investigated by Spearman's test.
<span custom-style = 'revision'>Differences in symptom frequency over time were investigated by Cochran Q test. 
Interaction effects were analyzed by two-way ANOVA with $\eta^2$ effect size statistic.</span>
<span custom-style = 'revision'>BIPQ dimensionality and consistency was investigated by factor analysis [@Bartlett1937] and McDonald's $\omega$ [@McDonald1999].</span>

Multi-parameter modeling was done with <span custom-style = 'revision'>regularized regression algorithms:</span> Elastic Net [@Zou2005; @Friedman2010], LASSO (least absolute shrinkage and selection operator) [@Tibshirani1996; @Friedman2010] and Bayesian LASSO [@Park2012; @Gramacy2022]. 
<span custom-style = 'revision'>The choice of regularized linear regression was motivated by multi-dimensional character of the study data set, i.e. comparable number of explanatory variables and observations [@Kirpich2018; @Morozova2015]. 
Ordinary least square models with backward term elimination developed in the study data set had poor reproducibility and over-parameterization as demonstrated by fit error expansion in cross-validation.</span>

<span custom-style = 'revision'>Modeling responses: the total IP score, emotion/concern/consequence and lacking control/coherence scores, were square root-transformed to guarantee normality.</span> 
Both first and second order terms of numeric explanatory variables were included in the models. 
Numeric explanatory and response variables were Z-score normalized. 
The optimal $\lambda$ for Elastic Net and LASSO were obtained by 200-repeats 10-fold cross-validation. 
The 'sparsity' parameter in Bayesian LASSO was found by 10-repeats 10-fold cross-validation [@Kuhn2008]. 
Explained variance ($R^2$) and model root mean squared error were assessed in the entire data set and 10-repeats 10-fold cross-validation [@Kuhn2008]. 
Elastic Net and LASSO model coefficients were calculated for the optimal $\lambda$ values. 
Bayesian LASSO coefficients were calculated as medians over all algorithm iterations [@Park2012]. 
<span custom-style = 'revision'>The number of non-zero model coefficients was found to differ substantially between the models as reported for data sets of similar size [@Morozova2015].</span>
<span custom-style = 'revision'>Hence,</span> variables with non-zero coefficients in <span custom-style = 'revision'>models constructed with all three algorithms</span> were deemed the key factors for 
<span custom-style = 'revision'>the total IP score, emotion/concern/consequences and lacking control/coherence scores.</span>

Clustering by the BIPQ items was accomplished by PAM (partitioning around medoids) <span custom-style = 'revision'>with Euclidean distance [@Schubert2019; @Drost2018].</span> 
<span custom-style = 'revision'>The clustering algorithm choice was motivated by its optimal stability in 10-fold cross-validation [@Lange2004] and good explanatory performance measured by the fraction of clustering variance as compared with hierarchical and KMEANS clustering.</span>

@@

The CovILD cohort was recruited <span custom-style = 'revision'>during the first outbreak of SARS-CoV-2</span> between April and June 2020 [@Sonnweber2020]. 
Out of 145 participants enrolled, `r nrow(cov_data$clear_data)` individuals with PSS, lung CT, LFT or cardiological abnormalities, <span custom-style = 'revision'>complete BIPQ and a complete set of study variables set at the one-year follow-up</span> were analyzed. 
The major reasons of patient dropout were missing follow-ups and incomplete BIPQ (__Figure \@ref(fig:fig-consort)__).
<span custom-style = 'revision'>The excluded participants tended towards less severe acute COVID-19, more dermatological and gastrointestinal symptoms at the one-year follow-up, and lower blood hemoglobin and ferritin as compared with the analysis collective; those effects were, however, not significant. 
The excluded participants had also significantly lower SMWD (__Supplementary Figure S3__).</span>

The <span custom-style = 'revision'>analyzed</span> participants were predominantly male (`r get_percent(cov_data$clear_data, "sex")['male']`%), the median age at COVID-19 diagnosis was `r get_median_iqr(cov_data$clear_data, "age")['median']` years (IQR: `r get_median_iqr(cov_data$clear_data, "age")['iqr']`), over one-third were active or ex-smokers (`r get_percent(cov_data$clear_data, "smoking_history")['yes']`%). 
Most participants suffered from comorbidities (`r get_percent(cov_data$clear_data, "comorb_present")['yes']`%), with cardiovascular disease, metabolic and respiratory disorders as leading conditions. 
The participants were classified by COVID-19 severity as ambulatory (`r get_percent(cov_data$clear_data, "cat_WHO")[1]`%), moderate (hospitalized, no intensive care, no oxygen therapy, `r get_percent(cov_data$clear_data, "cat_WHO")[2]`%) and severe COVID-19 survivors (hospitalized, oxygen therapy or intensive care, `r get_percent(cov_data$clear_data, "cat_WHO")[3]`%). 
The ambulatory COVID-19 subset had the lowest median age, smoking and comorbidity rates (__Table \@ref(tab:tab-cohort-baseline)__).
